---
title: "Reproducible Research: Peer Assessment 1"
output: html_document
---



## Downloading and preprocessing the data

```{r data, echo = TRUE}
download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",destfile = "./data.zip",method = "curl")

unzip("data.zip")

activity <- read.csv(file = "activity.csv",header = T,stringsAsFactors = F)
```

## What is mean total number of steps taken per day?

*   Histogram of the total number of steps taken each day

```{r 2, echo=TRUE}
steps_day<-aggregate(activity$steps, by = list(activity$date), FUN = sum, na.rm = TRUE)
colnames(steps_day) <- c("date","steps")

hist(steps_day$steps,col = "#20639B",xlab = "Total Steps",ylab = "Frequency",main = "Total Number of Steps per Day")
```

*   Mean and median number of steps taken each day

```{r 3, echo=TRUE}
mean(steps_day$steps)

median(steps_day$steps)
```

## What is the average daily activity pattern?

*   Time series plot of the interval and the average number of steps taken, averaged across all days

```{r 4, echo=TRUE}
steps_daily<-aggregate(activity$steps, by = list(activity$interval), FUN = mean, na.rm = TRUE)
colnames(steps_daily) <- c("interval","steps")

plot(steps_daily$interval,steps_daily$steps,type = "l",xlab = "Interval",ylab = "Average Steps",main = "Average Steps by interval averaged over day")
```

*   The 5-minute interval which contains the maximum number of steps is:
```{r 5,echo=TRUE}
match(max(steps_daily$steps),steps_daily$steps)

steps_daily[104,1]
```
The 5-minute interval that contains the maximum number of steps is 835.


## Imputing missing values
*   The total number of missing values in the dataset:
```{r 6, echo=TRUE}
  sum(is.na(activity$steps))
```

*   Strategy for filling in all of the missing values in the dataset: All the NA values have been filled by using the average(Which has been calculated in "steps_daily") values of steps corresponding to that interval.

```{r 7,echo=TRUE,warning=F}
index_NA <- which(activity$steps %in% NA)
for (i in 1:sum(is.na(activity$steps))) 
  {
  activity$steps[index_NA[i]]<-steps_daily[steps_daily==activity$interval[index_NA[i]],2]
}

#Check for NA values
sum(is.na(activity$steps))

#Creating new variable
steps_day_new <- aggregate(steps ~ date, activity, FUN = sum, na.rm = TRUE)

```
*   New histogram with replaced values
```{r 8,echo=T}
hist(steps_day_new$steps,col = "#F6D55C",xlab = "Total Steps",ylab = "Frequency",main = "Total Number of Steps per Day")

## Mean and median values are same after replacement and are different from the values obtained in the first part.

mean(steps_day_new$steps)

median(steps_day_new$steps)
```

## Are there differences in activity patterns between weekdays and weekends?

*   A new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r 9, echo=T}
 activity$date <- as.POSIXct(activity$date)
 activity$day <- sapply(activity$date, function(x) {
        if (weekdays(x) == "Saturday" | weekdays(x) =="Sunday") 
                {y <- "Weekend"} else 
                {y <- "Weekday"}
                y
        })
```

*   A panel plot containing a time series plot  of the interval and the average number of steps taken, averaged across all weekday days or weekends.

``` {r 10, echo = T}
steps_by_week <- aggregate(steps~interval + day, activity, mean, na.rm = TRUE)

library(lattice)

xyplot(steps ~ interval | day, groups = day, steps_by_week, outer = T, grid = T,type="l")

## The average steps over the weekends show higher pattern than that of the weekdays.
```